Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-04-28 stochastic_vehicle_dynamics.tex
[MM] adherence constraint adjustments

The base version of the constraint for each axle can be expressed as a function of $\bx$ and $\bu$, as follows:
\begin{equation}
	h^\textrm{A}_j(\bx,\bu) = \left[ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2\right]\frac{1}{Z_j^2(\bx,\bu)}-1\leq0\qquad j=1,2.\label{eq:adherence_constraint}
\end{equation}
where the superscript A denotes adherence, and the subscript $j$ indicates the axle index, with $j=1$ referring to the front axle and $j=2$ to the rear axle. Here, $X_j$ and $Y_j$ denote the longitudinal and lateral components of the in-plane ground forces, respectively, as illustrated in Figure~\ref{fig:vehicle_model}, while $Z_j$ represents the total vertical load acting on the axle. 
As can be seen from~\eqref{eq:adherence_constraint}, the points $\left(X_j,Y_j\right)$ are constrained to lie within an ellipse whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$.

The gradient of the constraint w.r.t. the state vector $\bx$, denoted by $\na_{\bx}h^\textrm{A}_j(\bmu)$, is used to evaluate the standard deviation $\sigma^\textrm{A}_j$ associated with the constraint. This gradient can be computed using the chain rule, as follows:
\begin{equation}
	\na_{\bx}h^\textrm{A}_j(\bmu,\bu) = 
	\frac{\pd h^\textrm{A}_j}{\pd X_j}\at_{\bmu,\bu}\na_{\bx}X_j(\bmu,\bu) + 
	\frac{\pd h^\textrm{A}_j}{\pd Y_j}\at_{\bmu,\bu}\na_{\bx}Y_j(\bmu,\bu) + 
	\frac{\pd h^\textrm{A}_j}{\pd Z_j}\at_{\bmu,\bu}\na_{\bx}Z_j(\bmu,\bu)
\end{equation}
where the first element is zero because $X_j$ is a function of the input vector $\bu$ only.

For example, introducing a back-off term $\be=0.1$ in the constraint implies that the axle's adherence capacity can be exploited up to 90\% of its limit. This allows the optimizer to determine the most appropriate trade-off between the longitudinal force $X$ and the lateral force $Y$ --- often by reducing both components to some extent --- based on the specific requirements of the manoeuvre.

>>>>

The base version of the constraint for each axle can be expressed as a function of $\bx$ and $\bu$, as follows:
\begin{equation}
	h^\textrm{A}_j(\bx,\bu) = \left[ \left( \frac{X_j(\bx,\bu)}{\mu_{x,j}} \right)^2+ \left( \frac{Y_j(\bx,\bu)}{\mu_{y,j}}\right)^2\right]\frac{1}{Z_j^2(\bx,\bu)}-1\leq0\qquad j=1,2.\label{eq:adherence_constraint}
\end{equation}
where the superscript A denotes adherence, and the subscript $j$ indicates the axle index, with $j=1$ referring to the front axle and $j=2$ to the rear axle. Here, $X_j$ and $Y_j$ denote the longitudinal and lateral components of the in-plane ground forces, respectively, as illustrated in Figure~\ref{fig:vehicle_model}, while $Z_j$ represents the total vertical load acting on the axle. 
As can be seen from~\eqref{eq:adherence_constraint}, the points $\left(X_j,Y_j\right)$ are constrained to lie within an ellipse whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$. The constraint, without the back-off term, is represented in the left panel of Figure~\ref{fig:robust_constraints} by the solid line ellipse in the plane $X_j-Y_j$.

To better clarify the effect of the back-off term on the constraint we can suppose to have a combination of $\bx$, $\bu$, $P$ and $\ga^\textrm{A}_j$ so that $\be^\textrm{A}_j=0.2$. This implies that the semi-axes of the ellipse are reduced by a factor $\sqrt{1-\be^\textrm{A}_j} = \sqrt{0.8}$, as shown in the left panel Figure~\ref{fig:robust_constraints}, where the ellipse in dashed line delimits the available region with the back-off applied. This allows the optimizer to determine the most appropriate trade-off between the longitudinal force $X_j$ and the lateral force $Y_j$ --- often by reducing both components to some extent --- based on the specific requirements of the manoeuvre.

%The gradient of the constraint w.r.t. the state vector $\bx$, denoted by $\na_{\bx}h^\textrm{A}_j(\bmu)$, is used to evaluate the standard deviation $\sigma^\textrm{A}_j$ associated with the constraint. This gradient can be computed using the chain rule, as follows:
%\begin{equation}
%	\na_{\bx}h^\textrm{A}_j(\bmu,\bu) = 
%	\frac{\pd h^\textrm{A}_j}{\pd X_j}\at_{\bmu,\bu}\na_{\bx}X_j(\bmu,\bu) + 
%	\frac{\pd h^\textrm{A}_j}{\pd Y_j}\at_{\bmu,\bu}\na_{\bx}Y_j(\bmu,\bu) + 
%	\frac{\pd h^\textrm{A}_j}{\pd Z_j}\at_{\bmu,\bu}\na_{\bx}Z_j(\bmu,\bu)
%\end{equation}
%where the first element is zero because $X_j$ is a function of the input vector $\bu$ only.

\begin{figure}
	\centering
	\begin{tikzpicture}[scale = .7]
		% Parameters
		\def\outerX{3}   % x radius of outer ellipse
		\def\outerY{2}   % y radius of outer ellipse
		\def\gap{0.2}    % distance between outer and inner ellipse
		\def\innerX{2.68}
		\def\innerY{1.79}

		% Draw outer ellipse
		\draw[] (0,0) ellipse [x radius=\outerX, y radius=\outerY];

		% Draw inner ellipse filled with color
		\fill[black!30] (0,0) ellipse [
		x radius={\innerX},
		y radius={\innerY}
		];
		\draw[dashed] (0,0) ellipse [
		x radius={\innerX},
		y radius={\innerY}
		];

		% Draw axes
		\draw[->] (-\outerX-1,0) -- (\outerX+1,0) node[right] {$X_j$};
		\draw[->] (0,-\outerY-1) -- (0,\outerY+1) node[above] {$Y_j$};

		% Arrows pointing to the ellipses
		\draw[->,thin] (1.7,2) -- ({\outerX*cos(60)},{\outerY*sin(60)}) node[midway, above right] {$\beta=0$};
		\draw[->,thin] (2.3,1.7) -- ({(\innerX)*cos(45)},{(\innerY)*sin(45)}) node[midway, right, shift={(0.05,.2)}] {$\beta=\gap$};

		% Arrows pointing to the ellipses
		\draw[->,thin] ({\outerX+.3},.3) -- (\outerX,0) node[midway, above right,shift={(.05,0)}] {$\mu_x Z_j$};
		\draw[->,thin] (.3,{\outerY+.3}) -- (0,\outerY) node[midway, above right,shift={(.05,0)}] {$\mu_y Z_j$};

	\end{tikzpicture}
	\hfill
	\begin{tikzpicture}[scale = .7]
		% Parameters
		\def\outerX{3}   % x radius of outer ellipse
		\def\outerY{2}   % y radius of outer ellipse
		\def\gap{0.2}    % distance between outer and inner ellipse
		\def\innerX{2.68}
		\def\innerY{1.79}

		% Draw outer ellipse
		\draw[] (0,0) ellipse [x radius=\outerX, y radius=\outerY];

		% Draw inner ellipse filled with color
		\fill[black!30] (0,0) ellipse [
		x radius={\innerX},
		y radius={\innerY}
		];
		\node[above] at (0,0) {PATH CONSTRAINT FIG};
		\draw[dashed] (0,0) ellipse [
		x radius={\innerX},
		y radius={\innerY}
		];

		% Draw axes
		\draw[->] (-\outerX-1,0) -- (\outerX+1,0) node[right] {$X_i$};
		\draw[->] (0,-\outerY-1) -- (0,\outerY+1) node[above] {$Y_i$};

		% Arrows pointing to the ellipses
		\draw[->,thin] (1.7,2) -- ({\outerX*cos(60)},{\outerY*sin(60)}) node[midway, above right] {$\beta=0$};
		\draw[->,thin] (2.3,1.7) -- ({(\innerX)*cos(45)},{(\innerY)*sin(45)}) node[midway, right, shift={(0.05,.2)}] {$\beta=\gap$};

		% Arrows pointing to the ellipses
		\draw[->,thin] ({\outerX+.3},.3) -- (\outerX,0) node[midway, above right,shift={(.05,0)}] {$\mu_x Z_i$};
		\draw[->,thin] (.3,{\outerY+.3}) -- (0,\outerY) node[midway, above right,shift={(.05,0)}] {$\mu_y Z_i$};

	\end{tikzpicture}
	\caption{Graphical representation of the two constraints analyzed in the section. The left panel shows the adherence constraint in the plane $X_j-Y_j$. The ellipse in solid line, whose semi-axes are given by $\mu_{x,j}Z_j$ and $\mu_{y,j}Z_j$, represents the border of the available region for the base constraint, without back-off. The ellipse in dashed line delimits the available region when a back-off $\be^\textrm{A}_j=0.2$ is present. This ellipse has its semi-axes reduced by a factor $\sqrt{1-\be^\textrm{A}_j}$ w.r.t. the one in solid line. 
	}
	\label{fig:robust_constraints}
\end{figure}
